---
title: "SpliceImpactR"
author:
- name: Zachary Wakefield
  affiliation:
  - &1 Boston University, Boston, MA
  - &2 Bioinformatics Program, Boston University, Boston, MA
- name: Ana Fiszbein
  affiliation:
  - *1
  - *2
date: '`r format(Sys.Date(), "%B %e, %Y")`'
package: SpliceImpactR
output:
    BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{SpliceImpactR}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options:
    chunk_output_type: console
---

```{r, include=FALSE, messages=FALSE, warnings=FALSE}
knitr::opts_chunk$set(comment="", fig.align="center", fig.width=8.75, cache=FALSE)
library(devtools)
library(tidyverse)
library(data.table)
library(dplyr)
library(SpliceImpactR)
```



# Introduction
Paragraph about package

# Documentation
Please visit <https://montilab.github.io/hypeR-docs/> for documentation, examples, and demos for all features and usage or read our recent paper [hypeR: An R Package for Geneset Enrichment Workflows](https://doi.org/10.1093/bioinformatics/btz700) published in _Bioinformatics_. 

# Requirements

We recommend the latest version of R (\>= 4.0.0) but **hypeR** currently requires R (\>= 3.6.0) to be installed directly from Github or Bioconductor. To install with R (\>= 3.5.0) see below. Use with R (\< 3.5.0) is not recommended.

# Installation

Install the development version of the package from Github.

``` r
devtools::install_github("fiszbein-lab/SpliceImpactR")
```

Or install the development version of the package from Bioconductor.

``` r
BiocManager::install("fiszbein-lab/SpliceImpactR", version="devel")
```

 Or install with  Conda.
```bash
conda create --name spliceimpactr
source activate spliceimpactr
conda install -c r r-devtools
R
library(devtools)
devtools::install_github("fiszbein-lab/SpliceImpactR")
```

Or install with previous versions of R.
```bash
git clone https://github.com/montilab/hypeR
nano SpliceImpactR/DESCRIPTION
# Change Line 8
# Depends: R (>= 3.6.0) -> Depends: R (>= 3.5.0)
R
install.packages("path/to/SpliceImpactR", repos=NULL, type="source")
```

# Usage
## Access gencode information
__SpliceImpactR__ requires the acceession of various genome annotations, accessed through biomaRt and directly through gencode, here we access the gencode files
```{r, eval=FALSE}
## To save time, we add a directory path to save the annotations
translations <- SpliceImpactR::getTranslations("/path/")

transcripts <- SpliceImpactR::getTranscripts("/path/")
```

## Access biomaRt information
Here we obtain further annotations through biomaRt
```{r, eval=FALSE}
## Getting access to biomaRt sometimes proves challenging, if errors in completing the ensembl call, run the following commented lines in the same chunk as the call
# new_config <- httr::config(ssl_verifypeer = FALSE)
# httr::set_config(new_config, override = FALSE)

ensembl <- biomaRt::useEnsembl(biomart = "ensembl",
                               dataset = "hsapiens_gene_ensembl")

## To save time, we add a directory path to save the annotations
biomart_data <- SpliceImpactR:::setupBiomart("/path/")

gtf <- SpliceImpactR::getAnnotation(biomart_data)
```

## Toy genome annotation loading
For the sake of this vignette, we use toy versions of the annotations for speed and ease
```{r}
# For purposes of these examples, data directory is in extdata, however this should be replaced with the directory path to  data
dataDirectory <- paste0(system.file("extdata", package="SpliceImpactR"), "/")

transcripts_sample <- list(transDF = readr::read_csv(paste0(dataDirectory, "transcripts_limited_transDF.csv"), 
                                                           show_col_types = FALSE),
                       c_trans = readr::read_lines(paste0(dataDirectory, "transcripts_limited_c_trans.csv")))

gtf_sample <- list(gtf = readr::read_csv(paste0(dataDirectory, "gtf_limited.csv"), 
                                                           show_col_types = FALSE),
            transcript_gtf = readr::read_csv(paste0(dataDirectory, "transcript_gtf_limited.csv"), 
                                                           show_col_types = FALSE),
            tgp_biomart = readr::read_csv(paste0(dataDirectory, "tgp_biomart_limited"), 
                                                           show_col_types = FALSE)
            )

translations_sample <- readr::read_lines(paste0(dataDirectory, "translations_limited.csv"))

biomart_data_sample <- list(ip = readr::read_csv(paste0(dataDirectory, "biomart_ip.csv"), 
                                                           show_col_types = FALSE),
                            code_regions = readr::read_csv(paste0(dataDirectory, "biomart_code_regions.csv"), 
                                                           show_col_types = FALSE),
                            pfam_exon_level = readr::read_csv(paste0(dataDirectory, "biomart_pfam_exon_level.csv"), 
                                                           show_col_types = FALSE),
                            fsd_exon_data = readr::read_csv(paste0(dataDirectory, "biomart_data_sample.csv"), 
                                                           show_col_types = FALSE),
                            pfam_data = readr::read_csv(paste0(dataDirectory, "biomart_pfam_exon.csv"), 
                                                           show_col_types = FALSE))

```


## Setting up data_df to identify conditions for comparison
test_group and control_group should be formatted as vectors of the path to the files from the HIT Index and/or rMATS -- without the ".{type}PSI" ending (eg: if the AFEPSI files is in /files/sample1.AFEPSI, the path given should be /files/sample1)
Rmats files should be renamed to be the same as the HIT Index file paths, just with the respective alternative splicing event: /files/sample1.SEPSI, /files/sample1.RIPSI
```{r}
# We are feeding the toy data in extdata/rawData here for computational speed
test_group <- paste0(dataDirectory, "rawData/", c("test1","test2", "test3"))
control_group <- paste0(dataDirectory, "rawData/", c("control1", "control2", "control3"))

# the data_df should be organized as such
data_df <- data.frame(
    sample_names = c(control_group, test_group),
    phenotype_names = c(
      rep("control", length(control_group)),
      rep("test", length(test_group))
     ),
   stringsAsFactors = FALSE
  )
```


## Basic Usage
The basic usage is able to be performed in one main call, fullASoutcome. THis generates output from multiple alternative RNA processing event types 
```{r}
# fullASoutcome performs the full pipeline on multiple alternative RNA processing events, only performed on AFE, SE, HIT here
twoASfullRun <- fullASoutcome(as_types = c("AFE", "SE", "HIT"),
                              output_directory = NULL,
                              data_directory = dataDirectory,
                              data_df,
                              outlier_handle = "Inf",
                              cutoff = .1,
                              cores = 1,
                              bg_pre = NA,
                              tti_location = NULL,
                              initTTI = NULL,
                              mOverlap = .05,
                              plotAlignments = FALSE,
                              s_gtf = gtf_sample,
                              transcripts = transcripts_sample,
                              translations = translations_sample,
                              biomart_data = biomart_data_sample,
                              max_zero_prop = 1,
                              min_prop_samples = 0,
                              chosen_method = 'qbGLM')

```


## Output walkthrough
For each alternative RNA processing event and the HIT Index comparison, we have various outputs
```{r}
twoASfullRun$AFE$pfg$gdf
```


Proximal shift plot
```{r}
twoASfullRun$AFE$proxPlot$proxPlot
```

Protein length comparison plot
```{r}
twoASfullRun$AFE$length_comparison
```

Overview comparison plot
```{r}
twoASfullRun$AFE$initial_comparison$comb_plot
```


## Integrated analysis
In order to perform the holistic analysis, we need multiple types of altenrative RNA processing events with valid outputs -- in the toy data, we just provide valid AFE changes for speed
```{r}
fg_list <- list("AFE" = twoASfullRun$AFE$fg$proBed)
pfg_list <- list("AFE" = twoASfullRun$AFE$pfg$paired_proBed)
domain_list <- list("AFE" = twoASfullRun$AFE$gD$data)

integrated <- getIntegratedResults(fg_list,
                                   pfg_list,
                                   domain_list)

integrated$plots
```
